Decoding UK Hospital Wait Times: Systemic Challenges Beyond Resources

### 2024-12-03 paper/paper.qmd
Completed idealized methodology and LLM usage.

The estimand in this study is the average percentage change from 2015 in wait times for fifteen key fields of medicine @tbl-treatments. The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. This measure quantifies how changes in healthcare infrastructure and demand impact system-wide delays, offering insights into the relative contribution of each predictor to observed wait times.

@@

@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of the 3 predictor variables and estimand and 8 observations spanning the years 2015 to 2022. The predictors include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. The outcome variable `Total` captures wait times (in percent change from 2015 levels) for key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery. All metrics are standardized based on population or healthcare utilization data for each year. The dataset including individual treatment waiting times percetentage change from 2015 is provided in @tbl-dataappendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.

@@

```{r}
#| echo: false
#| warning: false
#| error: false
#| label: fig-total
#| tbl-cap: "Percentage Increase in Total Wait Times Compared to 2015"

# Plot for the "Total" column
ggplot(cleaned_data , 
       aes(x = Year, y = Total)) +
  geom_line(size = 1, color = "black") +
  geom_point(size = 2, color = "black") +
  labs(
    title = "Percentage Change in Total Wait Times Compared to 2015",
    x = "Year",
    y = "Percentage Change (%)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 11)
  )

@@

### Hospital Beds per 1,000
```{r}
#| label: fig-beds
#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2019
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2019", x = "Year", y = "Beds per 1 000 Inhabitants") +
  theme_minimal()
```

@@

```{r}
#| label: fig-physicians
#| fig-cap: United Kingdom's Physician rate per 1000 Inhabitants from 2015 to 2019
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = Year, y = physicians_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Physicians per 1 000 rates from 2015 to 2019", x = "Year", y = "Physicians per 1 000") +
  theme_minimal()
```

@@

### Demand for Services 
```{r}
#| label: fig-attendance-demand
#| fig-cap: "Type 1 Major Care Attendance in the United Kingdom from 2015 to 2019"
#| echo: false
#| warning: false

# Plotting Type 1 Major Care Attendance
ggplot(cleaned_data) +
  geom_line(aes(x = Year, y = attendance), size = 1) +
  scale_color_manual(values = c("A&E Attendances" = "blue")) +
  labs(
    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2019)",
    x = "Year",
    y = "Percentage Change from 2015 (%)",
    color = "Service Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", text = element_text(size = 11))
```
@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2022. Between 2015 and 2019, A&E attendances showed a modest, steady increase, reflecting a growing demand for emergency services. However, there was a sharp decline in 2020, likely attributable to the COVID-19 pandemic and related restrictions, which disrupted normal healthcare-seeking behavior. Attendance levels rebounded significantly in 2021, surpassing pre-pandemic levels, and remained stable through 2022. This trend highlights both the resilience of demand for emergency services and the need for strategies to manage surges in attendance effectively, such as enhancing access to primary care and preventive healthcare services.

@@

# Model {#sec-model}
After conducting exploratory analysis on the dataset, we observed relationships between healthcare system factors (e.g., physicians per 1,000 people, hospital beds per 1,000 people, and Type 1 attendance rates) and wait times for medical procedures. These variables indicate potential predictive power, suggesting a linear relationship. To further investigate and quantify how these factors influence wait times, we implemented a linear regression model.

@@

* $Y_{i}$: The average percentage change in wait time for the 15 specific medical procedures in year $i$. 
* $X_{1i}, X_{2i}, X_{3i}$: The independent variables representing physicians per 1,000 people, beds per 1,000 people, and Type 1 attendance, respectively, for procedure $j$ in year $i$. 
* $\beta_0$: The constant term, representing the expected wait time when all predictors are zero.
* $\beta_1, \beta_2, \beta_3$: The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.
* $\epsilon_{i}$: The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.

@@

@@

This model framework allows us to evaluate the impact of key healthcare system factors on wait times. The selected predictors—beds_per_1000, physicians_per_1000, and attendance—are directly linked to resource availability and demand, making them relevant for understanding delays in medical care. The results from this model can provide actionable insights to inform policies aimed at optimizing resource allocation, improving healthcare infrastructure, and managing patient flow to reduce wait times. This analysis will help identify key leverage points for optimizing resource allocation and reducing wait times across various medical procedures.

@@

### Evaluation and Validation
Evaluation and Validation
To ensure the robustness of our findings, the model was validated through the following steps:

1. Residual Diagnostics: We assessed residual plots for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
2. Significance Testing: We conducted t-tests to assess whether each coefficient ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$) is statistically significant at the 0.05 level @sec-results.
3. Goodness of Fit: The R-squared statistic was calculated to measure how well the model explains the variability in wait times @sec-results.
4. Alternative Models: a variant containing data only from 2015 to 2019 was tested to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit @sec-adapted. In both models, the p-values of the coefficients were insignificant so the model containing the larger data set (original model from 2015-2022) was used for final analysis.

@@

```{r}
#| echo: false
#| warning: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors"

# Load the saved model
model <- readRDS("../models/wait_times_model.rds")

The linear regression model, presented in @tbl-modelresults, explores the relationships between healthcare system factors and the percentage change in wait times, though none of the variables demonstrate statistically significant effects, as indicated by p-values greater than 0.05. The number of hospital beds per 1,000 inhabitants shows a small positive effect ($\beta_1$=4.71), suggesting a minimal increase in wait times with higher bed availability, though this effect lacks statistical significance. Physician availability per 1,000 ($\beta_2$=31.83) indicates a larger positive association, yet it is also not statistically significant. Attendance rates ($\beta_3$ =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times. The model achieves an $R^2$ value of 0.599, indicating that approximately 60% of the variability in wait times is explained by the model, though the adjusted $R^2$ value of 0.299 reflects limited predictive power. With a sample size of n=8, the findings should be interpreted cautiously, as the lack of significance and low adjusted $R^2$ highlight potential limitations in the model's robustness and generalizability.

@@

```{r}
#| echo: false
#| warning: false
#| error: false
#| eval: true
#| label: fig-modelresults1
#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining percentage increase in wait times based on healthcare system predictors."

# Load the saved model
model <- readRDS("../models/wait_times_model.rds")

@@

  # Create the plot
  ggplot(data, aes_string(x = predictor_name, y = "Total")) +
    geom_point(color = "blue", size = 2, alpha = 0.7) +
    geom_line(data = pred_data, aes_string(x = predictor_name, y = "Predicted"), 
              color = "red", size = 1) +
    labs(
      x = predictor_name,
      y = "Wait Times (%)"
    ) +
    theme_minimal() +
    theme(aspect.ratio = 0.8)
}

# Combine the plots into one figure
combined_plot <- (p1 | p2 | p3) +
  plot_annotation(
    title = "Projections of Linear Model onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
  ) &
  theme(
    plot.title = element_text(size = 10)  # Adjust the title font size here
  )

@@

```
The projection plots in @tbl-modelresults illustrate the linear model's predictions for the percentage change in wait times (`Total`) compared to the three predictors: attendance, physicians_per_1000, and beds_per_1000. The red lines depict the predicted relationships, while the blue points represent the observed data values. The physicians_per_1000 predictor shows a clear positive relationship with wait times, as indicated by the upward-sloping red line, and most data points align reasonably well with the model's trend. Attendance also demonstrates a positive trend, but the variability in the observed data points around the red line suggests a weaker fit compared to attendance. In contrast, beds_per_1000 shows minimal association with wait times, with the red line nearly flat and the data points scattered without a clear trend. For each plot of the predictor vs outcome variable there seems to be at least two data points which deviate significantly from the remaining points. These were hypothesized to be due to the outliers from trend witnessed in @fig-total stemming from the Covid-19 pandemic years. @sec-adapted developed and analysed the significance of an adapted model for years 2015-2019 before the pandemic to avoid such outlier effects in capturing the trend. The overall model significance was much greater and linear assumption violations in the residual plots were non-existent, however individual coefficients were still non-significant and the dataset on which this model was constructed was even smaller with 5 observations.

@@

The positive but statistically insignificant relationship between physician density and wait times challenges the expectation that greater workforce availability reduces delays. This result may reflect systemic inefficiencies or increased demand driven by greater access to healthcare services. For example, higher physician availability can lead to more diagnostic and treatment referrals, overwhelming other parts of the healthcare system, such as specialized care or operating rooms. Additionally, unmeasured factors like disparities in physician specialization, workflow inefficiencies, or misalignment between physician density and patient needs could further obscure the relationship. These findings highlight that increasing physician numbers alone is insufficient to reduce wait times without addressing systemic bottlenecks. Interventions such as improving care pathways, enhancing interdisciplinary collaboration, and aligning physician availability with infrastructure and support staff could better leverage workforce capacity to reduce delays.

### Relationship between Wait Times and Type 1 Major Attendance
The weak positive association between attendance rates and wait times, though not significant, aligns with broader concerns about rising patient loads straining healthcare systems. The relationship is likely weak due to the significant deviation from pre-Covid trend in 2020, with a sudden drop by 25% from 2019 levels. This drop was recovered in 2021, but the impact on the linear model constructed by a limited 8 observation data set cannot be ignored. As attendance increases, hospitals may prioritize emergency cases, diverting resources from elective and non-urgent procedures, potentially exacerbating delays. Factors such as limited access to primary care, changes in population demographics, and healthcare-seeking behaviors likely drive the observed trends. Although the relationship was not statistically significant, addressing the root causes of rising attendance remains critical to reducing systemic strain. Potential strategies include expanding primary care access, implementing community triaging systems, and improving public health education to promote appropriate healthcare utilization. These measures could help balance patient loads and reduce delays for non-emergency procedures.

## Limitations 
The primary limitation of this study is the small sample size, which includes only eight observations spanning 2015 to 2022. This limits the generalizability of findings and weakens the statistical power needed to detect significant relationships. Additionally, the dataset is influenced by the effects of COVID-19, with wait times deviating significantly from the pre-pandemic trend. As illustrated in @fig-total, wait times followed a consistent linear increase from 2015 to 2019, but 2020 deviated sharply below trend, while 2021 exhibited a sharp increase above trend. These anomalies suggest inconsistent variance in the data and indicate that a linear model may not be appropriate for post-2019 data. Moreover, the dataset lacks information on key confounding variables such as regional healthcare policies, socioeconomic disparities, or detailed patient demographics, which could significantly influence wait times. Unmeasured systemic inefficiencies, such as staffing misalignment, operating room availability, or referral practices, may also account for the insignificant results. Furthermore, the linear model does not account for potential interaction effects or non-linear relationships between predictors, which could offer deeper insights into the dynamics of healthcare delays.

## Future Work
To address the limitations of this study, future research should prioritize collecting monthly data for the years 2015 to 2019 to explore more granular trends before the anomalies introduced by COVID-19. @sec-adapted shows the potential for this further investigation with a model consisting of the yearly data from 2015 to 2019 exhibiting a much stronger $R^2$ value of 0.875. This model is small (5 observations), thus a more extensive model involving the monthly data would be beneficial. Analyzing these pre-pandemic patterns could provide a clearer understanding of the role that metrics such as beds per 1,000, physician density, and attendance rates played during a relatively stable period. Additionally, expanding the dataset to include broader healthcare metrics, such as regional funding levels, technological investments, and patient demographics, could yield a more comprehensive analysis. Future studies should also explore non-linear models or interaction effects to capture the intricate relationships between predictors. Simulation models and scenario analyses could be used to test policy interventions, such as optimizing resource allocation or improving workflow efficiency, to identify actionable strategies for reducing wait times and improving healthcare delivery.

@@

## Full Analysis Dataset

@@

# Import the cleaned healthcare dataset
cleaned_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

@@

# Split and format the table into smaller parts
part1 <- cleaned_data[, c(1:6)] %>% format_numeric()
part2 <- cleaned_data[, c(6:12)] %>% format_numeric()
part3 <- cleaned_data[, c(12:18)] %>% format_numeric()
part4 <- cleaned_data[, c(18:ncol(cleaned_data))] %>% format_numeric()

@@

### Original Model Residuals {#sec-residuals}
```{r}
#| echo: false
#| warning: false
#| error: false
#| eval: true
#| label: fig-modeldetails
#| tbl-cap: "Residual plots of each predictor."

### Adapted Model {#sec-adapted}

```{r}
#| echo: false
#| output: asis
#| warning: false
#| eval: true
#| label: tbl-adj
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"

#### Read data ####
# Load the cleaned data
analysis_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

#### Filter data ####
# Include only data from 2015 to 2019
filtered_data <- analysis_data %>% 
  filter(Year >= 2015 & Year <= 2019)

#### Fit the linear model ####
# Model: Percentage change in wait times as a function of healthcare predictors
wait_times_model_adj <- lm(
  Total ~ beds_per_1000 + physicians_per_1000 + attendance,
  data = filtered_data
)

#### Save the model ####
# Save the fitted model for reuse
saveRDS(wait_times_model_adj, file = "../models/wait_times_model_pre_2020.rds")

# Display the model summary in a table
modelsummary::modelsummary(
  list("Frequentist Linear Model" = modela),
  statistic = c("p.value")
)
```
The linear regression model, summarized in @tbl-adj, examines the impact of healthcare system factors on wait times from 2015–2019. Hospital beds per 1,000 inhabitants ($\beta_1 =−176.99$) and physicians per 1,000 ($\beta_2 =−78.19$) both show negative associations with wait times, suggesting increased resources reduce delays, though neither effect is statistically significant ($p >0.05$). Attendance rates ($\beta_3=−0.66$) exhibit a minimal effect. The model explains 97.8% of the variability in wait times ($R^2$=0.978), though the adjusted $R^2$ of 0.911 reflects possible overfitting due to the small sample size ($n = 5$). While the model fits well, the lack of significant predictors highlights the need for further data to better capture drivers of wait times. The small RMSE of 0.58 indicates the linear trend is never far off from the actual wait times.

```{r}
#| echo: false
#| warning: false
#| eval: true
#| label: fig-residadj
#| tbl-cap: "Linear model explaining percentage increase in wait times based on healthcare system predictors (2015-2019)"
# Extract residuals and fitted values from the model
filtered_data <- filtered_data %>%
  mutate(
    residuals = residuals(modela),
    fitted = fitted.values(modela)
  )

@@

@@

  # Create the plot
  ggplot(data, aes_string(x = predictor_name, y = "Total")) +
    geom_point(color = "blue", size = 2, alpha = 0.7) +
    geom_line(data = pred_data, aes_string(x = predictor_name, y = "Predicted"), 
              color = "red", size = 1) +
    labs(
      x = predictor_name,
      y = "Wait Times (%)"
    ) +
    theme_minimal() +
    theme(aspect.ratio = 0.8)
}

# Combine the plots into one figure
combined_plot <- (p1 | p2 | p3) +
  plot_annotation(
    title = "Projections of Linear Model (2015-2019) onto \nEach Predictor vs Percentage Change in Total Wait Times Compared to 2015",
    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
  ) &
  theme(
    plot.title = element_text(size = 10)  # Adjust the title font size here
  )

>>>>

The estimand in this study is the is the median number of weeks spent waiting for treatment since referral averaged over fifteen key fields of medicine @tbl-treatments. The estimand will be modeled as a function of three predictors: hospital beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. This measure quantifies how changes in healthcare infrastructure and demand impact system-wide delays, offering insights into the relative contribution of each predictor to observed wait times.

@@

@tbl-data presents United Kingdom hospital and healthcare statistics from the cleaned dataset, consisting of the 3 predictor variables and estimand and 8 observations spanning the years 2015 to 2022. The predictors include year, beds per 1,000 inhabitants, physicians per 1,000 inhabitants, and Type 1 major care attendance rates. The outcome variable `Total` captures wait times (median number of weeks per year) averaged over key treatment areas such as cardiology, ENT, gynaecology, plastic surgery and neurosurgery. All metrics are standardized based on population or healthcare utilization data for each year. The dataset including individual treatment waiting times in weeks is provided in @tbl-dataappendix and the full list of treatments from which `Total` is computed is provided in @tbl-treatments.

@@

```{r}
#| echo: false
#| warning: false
#| error: false
#| label: fig-total
#| tbl-cap: "Median Number of Weeks for Total Wait Times Compared to 2015"

# Plot for the "Total" column
ggplot(cleaned_data , 
       aes(x = Year, y = Total)) +
  geom_line(size = 1, color = "black") +
  geom_point(size = 2, color = "black") +
  labs(
    title = "Median Number of Weeks for Total Wait Times Compared to 2015",
    x = "Year",
    y = "Average Wait Time (Median # of Weeks)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    text = element_text(size = 10),
    plot.title = element_text(size = 11),  
  )

@@

### Hospital Beds per 1,000
```{r}
#| label: fig-beds
#| fig-cap: United Kingdom's Hospital Beds per 1000 Inhabitants from 2015 to 2022
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = Year, y = beds_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Hospital Beds per 1 000 rates from 2015 to 2022", x = "Year", y = "Beds per 1 000 Inhabitants") +
  theme_minimal()
```

@@

```{r}
#| label: fig-physicians
#| fig-cap: United Kingdom's Physician rate per 1000 Inhabitants from 2015 to 2022
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = Year, y = physicians_per_1000)) +
  geom_line(color = "blue") +
  labs(title = "Physicians per 1 000 rates from 2015 to 2022", x = "Year", y = "Physicians per 1 000") +
  theme_minimal()
```

@@

### Demand for Services 
```{r}
#| label: fig-attendance-demand
#| fig-cap: "Type 1 Major Care Attendance in the United Kingdom from 2015 to 2022"
#| echo: false
#| warning: false

# Plotting Type 1 Major Care Attendance
ggplot(cleaned_data) +
  geom_line(aes(x = Year, y = attendance), size = 1) +
  scale_color_manual(values = c("A&E Attendances" = "blue")) +
  labs(
    title = "Type 1 Major Care Attendance in the United Kingdom (2015–2022)",
    x = "Year",
    y = "Percentage Change from 2015 (%)",
    color = "Service Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", text = element_text(size = 10), plot.title = element_text(size = 11),  ) 
```
@fig-attendance-demand illustrates the percentage change in Type 1 Major Care Attendance (A&E Attendances) in the UK from 2015 to 2022. Between 2015 and 2019, A&E attendances showed a modest, steady increase, reflecting a growing demand for emergency services. However, there was a sharp decline in 2020, likely attributable to the COVID-19 pandemic and related restrictions, which disrupted normal healthcare-seeking behavior. Attendance levels rebounded significantly in 2021, surpassing pre-pandemic levels, and remained stable through 2022. This trend highlights both the resilience of demand for emergency services and the need for strategies to manage surges in attendance effectively, such as enhancing access to primary care and preventive healthcare services.

@@

# Model {#sec-model}
After conducting exploratory analysis on the dataset, linear trends were observed in each of the predictors and the outcome variable for years 2015 to 2019. Years 2020 to 2022 exhibited significant deviations from the previous years' linear trend, likely due to the Covid19 pandemic. Thus years 2015 to 2019 were thought to be the best to model. Linear relationships were observed between healthcare system factors ( physicians per 1,000 people, hospital beds per 1,000 people, and Type 1 attendance rates) and wait times for medical procedures. These variables indicate potential predictive power, suggesting a linear relationship. To further investigate and quantify how these factors influence wait times, we implemented a multivariate linear regression model.

@@

* $Y_{i}$: The median number of weeks averaged over the 15 specific medical procedures in year $i$. 
* $X_{1i}, X_{2i}, X_{3i}$: The independent variables representing physicians per 1,000 people, beds per 1,000 people, and Type 1 attendance, respectively, for procedure $j$ in year $i$. 
* $\beta_0$: The constant term, representing the expected wait time when all predictors are zero.
* $\beta_1, \beta_2, \beta_3$: The slope coefficients for the predictors, representing the estimated change in wait time for a one-unit increase in the respective predictor.
* $\epsilon_{i}$: The error term, representing the deviation of the actual wait time from the predicted wait time based on the regression equation.

@@

These are checked in @sec-residuals.

@@

This model framework allows us to evaluate the impact of key healthcare system factors on wait times. The selected predictors $beds\_per\_1000$, $physicians\_per\_1000$, and $attendance$ are directly linked to resource availability and demand, making them relevant for understanding delays in medical care. The results from this model can provide actionable insights to inform policies aimed at optimizing resource allocation, improving healthcare infrastructure, and managing patient flow to reduce wait times. This analysis will help identify key leverage points for optimizing resource allocation and reducing wait times across various medical procedures.

@@

### Evaluation and Validation
To ensure the robustness of the findings, the model was validated through the following steps:

1. Residual Diagnostics: We assessed residual plots for patterns suggesting non-linearity or heteroscedasticity. No significant violations of assumptions were detected @sec-residuals.
2. Significance Testing: We conducted t-tests to assess whether each coefficient ($\beta_0$, $\beta_1$, $\beta_2$, $\beta_3$) is statistically significant at the 0.05 level @sec-results.
3. Goodness of Fit: The R-squared statistic was calculated to measure how well the model explains the variability in wait times @sec-results.
4. Alternative Models: a variant containing data from 2015 to 2022 was tested in @sec-adapted to confirm whether the anomalies present in the Covid-19 pandemic years of 2020-2021 was a probable cause for poor fit. This model had a much smaller $R^2$ and adjusted $R^2$ values and a much larger RMSE. The observation size differences between the two models were not significant enough to account for the much poorer fit, thus the smaller model was kept for final analysis.

@@

```{r}
#| echo: false
#| output: asis
#| warning: false
#| eval: true
#| label: tbl-adj
#| tbl-cap: "Linear model explaining wait times (median number of weeks per year) based on healthcare system predictors (2015-2019)"

# Display the model summary in a table
modelsummary::modelsummary(
  list("Multivariate Linear Model" = modela),
  statistic = c("p.value")
)
```
The linear regression model, summarized in @tbl-adj, evaluates the impact of healthcare system factors on wait times from 2015–2019, revealing that hospital beds per 1,000 inhabitants ($\beta_1 = -16.826$) and physicians per 1,000 inhabitants ($\beta_2 = -7.196$) are negatively associated with wait times, indicating that increased resources may reduce delays however the findings are statistically insignificant ($p > 0.05$), highlighting the limitations imposed by the small sample size ($n = 5$). Attendance rates ($\beta_3 = -0.075$) exhibit a negligible effect, with a low p-value (0.714) further suggesting minimal impact on wait times when controlling for other factors. The model explains 96.6% of the variability in wait times ($R^2 = 0.966$), but the adjusted $R^2$ of 0.866 points to potential overfitting, as the inclusion of weak predictors may inflate the apparent fit. A small root mean squared error (RMSE) of approximately 0.07 indicates that the model's predictions closely align with observed values, yet this performance may not generalize beyond the dataset. Despite the excellent overall fit, the lack of significant predictors underscores the limitations of the analysis, with the small sample size undermining the precision of coefficient estimates and the robustness of conclusions. Furthermore, the variability in predictor effects suggests the potential presence of omitted variables or non-linear interactions.

@@

@@

  # Create the plot
  ggplot(data, aes_string(x = predictor_name, y = "Total")) +
    geom_point(color = "blue", size = 2, alpha = 0.7) +
    geom_line(data = pred_data, aes_string(x = predictor_name, y = "Predicted"), 
              color = "red", size = 1) +
    labs(
      x = predictor_name,
      y = "Wait Times (# Weeks)"
    ) +
    theme_minimal() +
    theme(aspect.ratio = 0.8)
}

analysis_data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

# Include only data from 2015 to 2019
filtered_data <- analysis_data %>% 
  filter(Year >= 2015 & Year <= 2019)

# Combine the plots into one figure
combined_plot <- (p1 | p2 | p3) +
  plot_annotation(
    title = "Projections of Linear Model (2015-2019) onto \nEach Predictor vs Average Wait Times (Weeks)",
    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
  ) &
  theme(
    plot.title = element_text(size = 11),  
    axis.title.y = element_text(size = 9)
  )

@@

```
The projection plots in @fig-adjmodelresults illustrate the linear model's predictions for the average wait timein relation to the three predictors: attendance, physicians per 1,000, and beds per 1,000. The red lines in each plot represent the predicted relationships, while the blue points reflect the observed data. For attendance, the model suggests a negative relationship with wait times, as indicated by the downward-sloping red line. However, the observed data points display considerable variability around the prediction, with notable deviations suggesting the presence of unmodeled interactions or confounding factors. The relationship for physicians per 1,000 is also negative, with a steep downward slope showing that higher physician density is associated with reduced wait times. While the data points align more closely with the model's predictions compared to attendance, some deviations, particularly at lower physician densities, highlight potential limitations in the model’s assumptions. The strongest and most consistent relationship is observed for beds per 1,000, where the steep downward-sloping red line indicates a significant association between increased bed availability and reduced wait times. The observed data points align closely with this trend, suggesting that this predictor has a robust and straightforward influence on wait times.

Across all predictors, however, the small sample size (n = 5) significantly limits the reliability and generalizability of the findings. While the data for 2015–2019 avoids the confounding impact of the COVID-19 pandemic, the limited dataset reduces the ability to draw definitive conclusions, and individual coefficient estimates remain non-significant despite the improved model fit. The variability around the predictions, particularly for attendance and physicians per 1,000, highlights potential missing interactions or non-linear relationships that the linear model cannot capture. The model overall captures the relationship between the predictors and the outcome variable better than the model investigated in @sec-adapted.

@@

The negative but statistically insignificant relationship between physician density and wait times challenges the expectation that greater workforce availability reduces delays. This result may reflect systemic inefficiencies or increased demand driven by greater access to healthcare services. For example, higher physician availability can lead to more diagnostic and treatment referrals, overwhelming other parts of the healthcare system, such as specialized care or operating rooms. Additionally, unmeasured factors like disparities in physician specialization, workflow inefficiencies, or misalignment between physician density and patient needs could further obscure the relationship. These findings highlight that increasing physician numbers alone is insufficient to reduce wait times without addressing systemic bottlenecks. Interventions such as improving care pathways, enhancing interdisciplinary collaboration, and aligning physician availability with infrastructure and support staff could better leverage workforce capacity to reduce delays.

### Relationship between Wait Times and Type 1 Major Attendance
The weak negative association between attendance rates and wait times, though not significant, aligns with broader concerns about rising patient loads straining healthcare systems. The relationship is likely weak due to the significant deviation from pre-Covid trend in 2020, with a sudden drop by 25% from 2019 levels. This drop was recovered in 2021, but the impact on the linear model constructed by a limited 8 observation data set cannot be ignored. As attendance increases, hospitals may prioritize emergency cases, diverting resources from elective and non-urgent procedures, potentially exacerbating delays. Factors such as limited access to primary care, changes in population demographics, and healthcare-seeking behaviors likely drive the observed trends. Although the relationship was not statistically significant, addressing the root causes of rising attendance remains critical to reducing systemic strain. Potential strategies include expanding primary care access, implementing community triaging systems, and improving public health education to promote appropriate healthcare utilization. These measures could help balance patient loads and reduce delays for non-emergency procedures.

## Limitations 
The primary limitation of this study is the small sample size ($n = 5$ for the focused 2015–2019 model), which weakens the statistical power and limits the generalizability of the findings. The dataset also includes anomalies introduced by the COVID-19 pandemic, with wait times deviating sharply from the pre-pandemic trend in 2020 and 2021. These anomalies introduce inconsistent variance in the data, further reducing the reliability of the adapted linear model in @sec-adapted, which assumes consistent patterns across the dataset. As illustrated in @fig-total, wait times followed a consistent linear increase from 2015 to 2019, but 2020 deviated sharply below trend, while 2021 exhibited a sharp increase above trend. These anomalies suggest inconsistent variance in the data and indicate that a linear model may not be appropriate for post-2019 data. Moreover, the dataset lacks information on key confounding variables such as regional healthcare policies, socioeconomic disparities, or detailed patient demographics, which could significantly influence wait times. Unmeasured systemic inefficiencies, such as staffing misalignment, operating room availability, or referral practices, may also account for the insignificant results. Furthermore, the linear model does not account for potential interaction effects or non-linear relationships between predictors, which could offer deeper insights into the dynamics of healthcare delays.

## Future Work
To address the limitations of this study, future research should prioritize collecting monthly data for the years 2015 to 2019 to explore more granular trends before the anomalies introduced by COVID-19. @sec-results shows the potential for this further investigation with a model consisting of the yearly data from 2015 to 2019 exhibiting a much stronger $R^2$ value of 0.966 than the model in @sec-adapted. This model is small (5 observations), thus a more extensive model involving the monthly data would be necessary. Analyzing these pre-pandemic patterns could provide a clearer understanding of the role that metrics such as beds per 1,000, physician density, and attendance rates played during a relatively stable period. Additionally, expanding the dataset to include broader healthcare metrics, such as regional funding levels, technological investments, and patient demographics, could yield a more comprehensive analysis. Future studies should also explore non-linear models or interaction effects to capture the intricate relationships between predictors. Simulation models and scenario analyses could be used to test policy interventions, such as optimizing resource allocation or improving workflow efficiency, to identify actionable strategies for reducing wait times and improving healthcare delivery.

@@

## Analysis Dataset

@@

# Import the cleaned healthcare dataset
cleaned_data <- read_parquet("../data/02-analysis_data/healthcare_and_individual_treatments.parquet")

@@

# Split and format the table into smaller parts
part1 <- cleaned_data[, c(1:5)] %>% format_numeric()
part2 <- cleaned_data[, c(6:11)] %>% format_numeric()
part3 <- cleaned_data[, c(12:17)] %>% format_numeric()
part4 <- cleaned_data[, c(18:ncol(cleaned_data))] %>% format_numeric()

@@

### Original Model Residuals {#sec-residuals}

```{r}
#| echo: false
#| warning: false
#| eval: true
#| label: fig-residadj
#| tbl-cap: "Linear model explaining change in wait times based on healthcare system predictors (2015-2019)"
# Extract residuals and fitted values from the model
filtered_data <- filtered_data %>%
  mutate(
    residuals = residuals(modela),
    fitted = fitted.values(modela)
  )

@@

### Adapted Model {#sec-adapted}
```{r}
#| echo: false
#| warning: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Linear model explaining change in wait times based on healthcare system predictors"

data <- read_parquet("../data/02-analysis_data/final_healthcare_data.parquet")

# Load the saved model
model <- readRDS("../models/wait_times_model_2015_to_2022.rds")

The linear regression model, presented in @tbl-modelresults, explores the relationships between healthcare system factors and the percentage change in wait times, though none of the variables demonstrate statistically significant effects, as indicated by p-values greater than 0.05. The number of hospital beds per 1,000 inhabitants shows a small negative effect ($\beta_1$=-0.556), suggesting a minimal decrease in wait times with higher bed availability, though this effect lacks statistical significance. Physician availability per 1,000 ($\beta_2$=2.517) indicates a larger positive association, yet it is also not statistically significant. Attendance rates ($\beta_3$ =0.40) exhibit the smallest positive effect, suggesting a weak association with wait times. The model achieves an $R^2$ value of 0.599, indicating that approximately 60% of the variability in wait times is explained by the model, though the adjusted $R^2$ value of 0.299 reflects limited predictive power. With a sample size of n=8, the findings should be interpreted cautiously, as the lack of significance and low adjusted $R^2$ highlight potential limitations in the model's robustness and generalizability.

```{r}
#| echo: false
#| warning: false
#| error: false
#| eval: true
#| label: fig-modeldetails
#| tbl-cap: "Residual plots of each predictor."

# Add QQ Plot
plot_qq <- ggplot(data, aes(sample = residuals)) +
  stat_qq(color = "blue") +
  stat_qq_line(color = "red", linetype = "dashed") +
  theme_classic() +
  labs(
    title = "QQ Plot of Residuals",
    y = "Sample Quantiles",
    x = "Theoretical Quantiles"
  ) +
  theme(plot.title = element_text(size = 10))

```{r}
#| echo: false
#| warning: false
#| error: false
#| eval: true
#| label: fig-modelresults1
#| tbl-cap: "Projections of Linear Model onto Each Predictor: Linear model explaining chane in wait times based on healthcare system predictors."

# Load the saved model
model <- readRDS("../models/wait_times_model_2015_to_2022.rds")

@@

  # Create the plot
  ggplot(data, aes_string(x = predictor_name, y = "Total")) +
    geom_point(color = "blue", size = 2, alpha = 0.7) +
    geom_line(data = pred_data, aes_string(x = predictor_name, y = "Predicted"), 
              color = "red", size = 1) +
    labs(
      x = predictor_name,
      y = "Wait Times (#Weeks)"
    ) +
    theme_minimal() +
    theme(aspect.ratio = 0.8)
}

# Combine the plots into one figure
combined_plot <- (p1 | p2 | p3) +
  plot_annotation(
    title = "Projections of Linear Model onto \nEach Predictor vs Average Wait Times Compared to 2015 ",
    caption = "**Red lines indicate the model's predictions, and blue points represent the actual data."
  ) &
  theme(
    plot.title = element_text(size = 11),  
    axis.title.y = element_text(size = 9)
  )
